Enforced Invariant Confluence

### 2018-04-05 doc/paper/sections/introduction.tex
Finished first cut of abstract.
Documented segmented invariant-confluence DP.

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes, sometimes. Bailis et al.\
introduced the notion of invariant-confluence as a necessary and sufficient
condition for when invariants can be maintained over replicated data without
the need for any coordination~\cite{bailis2014coordination}. If an object is
invariant-confluent with respect to an invariant, we can replicate it with the
performance benefits of weak consistency and (some of) the correctness benefits
of strong consistency. However, the task of identifying whether or not an
object actually is invariant-confluent remain a challenge. Bailis et al.\
manually categorized a set of common objects, transactions, and invariants
(e.g.\ foreign key constraints on relations, linear constraints on integers) as
invariant-confluent or not, but ideally we would have a general purpose program
that can automatically determine invariant-confluence for us. \textbf{The main
thrust of this paper is to make progress towards a general-purpose
invariant-confluence decision procedure.}

@@

Next, we take a step beyond invariant-confluence and develop a generalization
of invariant-confluence called segmented invariant-confluence. While
invariant-confluence characterizes objects which can be replicated
\emph{without any} coordination, segmented invariant-confluence allows us to
replicate non-invariant-confluent objects with \emph{minimal coordination}. The
main idea is that invariant-satisfying states are divided into a number of
segments, and the set of allowable of transactions within each segment is
restricted. Within a segment, servers act without any coordination,
synchronizing only to transition across segment boundaries. This design
highlights the trade-off between application complexity and
coordination-freedom; more complex applications require more segments which
require more coordination, and vice-versa.

>>>>

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes, sometimes. Bailis et al.\
introduced the notion of invariant-confluence as a necessary and sufficient
condition for when invariants can be maintained over replicated data without
the need for any coordination~\cite{bailis2014coordination}. If an object is
invariant-confluent with respect to an invariant, we can replicate it with the
performance benefits of weak consistency and (some of) the correctness benefits
of strong consistency. However, the task of identifying whether or not an
object actually is invariant-confluent remain a challenge. Bailis et al.\
manually categorized a set of common objects, transactions, and invariants
(e.g.\ foreign key constraints on relations, linear constraints on integers) as
invariant-confluent or not, but ideally we would have a general purpose program
that can automatically determine invariant-confluence for us. \textbf{The first
main thrust of this paper is to make progress towards a general-purpose
invariant-confluence decision procedure.}

@@

\textbf{The second main thrust of this paper is to take a step beyond
invariant-confluence and develop a generalization of invariant-confluence
called segmented invariant-confluence.} While invariant-confluence
characterizes objects which can be replicated \emph{without any} coordination,
segmented invariant-confluence allows us to replicate non-invariant-confluent
objects with \emph{minimal coordination}. The main idea is that
invariant-satisfying states are divided into a number of segments, and the set
of allowable of transactions within each segment is restricted. Within a
segment, servers act without any coordination, synchronizing only to transition
across segment boundaries. This design highlights the trade-off between
application complexity and coordination-freedom; more complex applications
require more segments which require more coordination, and vice-versa.

### 2018-04-05 doc/paper/sections/segmented_invariant_confluence.tex
Documented segmented invariant-confluence DP.

\section{Segmented Invariant-Confluence}\seclabel{SegmentedInvariantConfluence}
If a distributed object is invariant-confluent, then the object can be
replicated without the need for any form of coordination to maintain the
object's invariant. But what if the object is \emph{not} invariant-confluent?
In this section, we present a generalization of invariant-confluence called
\defword{segmented invariant-confluence} that can be used to maintain the
invariants of non-invariant-confluent objects, requiring only a small amount of
coordination.

The main idea behind segmented invariant-confluence is to segment the state
space into a number of segments and restrict the set of allowable transactions
within each segment in such a way that the object is invariant-confluent
\emph{within each segment} (even though it may not be globally
invariant-confluent). Then, servers can run coordination-free within a segment
and need only coordinate when transitioning from one segment to another. We now
formalize this idea by defining segmented invariant-confluence and a new system
model to execute segmented invariant-confluent objects.

\subsection{Formalism}
Consider a distributed object $O = (S, \join)$, a start state $s_0 \in S$, a
set of transitions $T$, and an invariant $I$. A segmentation $\Sigma = (I_1,
T_1), \ldots, (I_n, T_n)$ is a sequence (not a set) of $n$ segments$(I_i, T_i)$
where every $T_i$ is a subset of $T$. $O$ is \defword{segmented
invariant-confluent} with respect to $s_0$, $T$, $I$, and $\Sigma$, abbreviated
\defword{\sTISconfluent}, if the following conditions hold:
\begin{itemize}
  \item
    The start state satisfies the invariant (i.e. $I(s_0)$).

@@

  \item
    For every $(I_i, T_i) \in \Sigma$ and for every state $s \in I_i$, $O$ is
    \sticonfluent{s}{T_i}{I_i}.
\end{itemize}

@@

\subsection{System Model}
Now, we describe a new system model used to replicate a segmented
invariant-confluent object with coordination-free execution within a segment
and a small amount of coordination to transition between segments. As before,
we replicate an object $O$ across a set $p_1, \ldots, p_n$ of $n$ servers each
of which manages a replica $s_i \in S$ of the object. Every server begins with
$s_0$, $T$, $I$, and $\Sigma$. Moreover, at any given point in time, a server
designates one of the segments in $\Sigma$ as its \defword{active segment}.
Initially, every server chooses the first segment $(I_i, T_i) \in \Sigma$ such
that $I_i(s_0)$ to be its active segment. We'll see momentarily the
significance of the active segment.

As before, servers repeatedly perform one of two actions: execute a transaction
or merge states with another server. Merging states in the new system model is
identical to merging states in the old system model. A server $p_i$ sends its
state $s_i$ to another server $p_j$ which \emph{must} merge $s_i$ into its
state $s_j$.
%
Transaction execution in the new system model, on the other hand, is a bit more
involved. Consider a server $s_i$ with active segment $(I_i, T_i)$. A client
can request that $p_i$ execute a transaction $t \in T$. We consider what
happens when $t \in T_i$ and $t \notin T_i$ separately.

@@

\begin{algorithm}[t]
  \caption{Transaction execution in new system model}%
  \algolabel{TxnExecution}
  \begin{algorithmic}
    \If{$t \notin T_i$}
      \State Execute $t$ with global coordination
    \Else{}
      \If{$I_i(t(s_i))$} Commit $t$
      \ElsIf{$\lnot I(t(s_i))$} Abort $t$
      \Else{} Execute $t$ with global coordination
      \EndIf{}
    \EndIf{}
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Discussion}
There are a couple of things worth noting about segmented invariant-confluence
and its system model. First, invariant-confluence is a very coarse-grained
property. If an object is invariant-confluent, then we can replicate it with no
coordination. If it is not invariant-confluent, then we have no guarantees.
There's no in-between. Segmented invariant-confluence, on the other hand, is a
much more fine-grained property that can be applied to applications with
varying degrees of complexity.  Segmented invariant-confluence provides
guarantees to complex applications that require a large number of segments and
to simple applications with a smaller number of segments, whereas
invariant-confluence only provides guarantees to applications that can be
segmented into a single segment.

Second, segmented invariant-confluence can naturally model distributed locking.
One approach to replicating non-invariant-confluent objects is to recognize
which transactions cannot be safely executed concurrently and then require them
to acquire a distributed lock before executing~\cite{balegas2015putting,
gotsman2016cause}. For example, in a banking application with the invariant
that all balances remain positive, concurrent deposits are permitted, but
concurrent withdrawals can lead to invariant violations, thus forcing
withdrawals to first acquire a distributed lock before executing. To model this
locking with segmented invariant-confluence, we simply remove lock-acquiring
transactions from every segment's set of transactions. Consequently, executing
these transactions are executed with global coordination, achieving the same
effect as acquiring a distributed lock. In fact, segmented invariant-confluence
can be used to model more fine-grained locking. Omitting a transaction from
only certain segments is analogous to requiring a transaction to acquire a
distributed lock only when certain conditions are met.

% TODO: Forward reference examples in the evaluation section.

@@

>>>>

\section{Segmented Invariant-Confluence}\seclabel{SegmentedInvariantConfluence}
\newcommand{\IsIclosed}{\textsf{IsIclosed}}

If a distributed object is invariant-confluent, then the object can be
replicated without the need for any form of coordination to maintain the
object's invariant. But what if the object is \emph{not} invariant-confluent?
In this section, we present a generalization of invariant-confluence called
\defword{segmented invariant-confluence} that can be used to maintain the
invariants of non-invariant-confluent objects, requiring only a small amount of
coordination.

The main idea behind segmented invariant-confluence is to segment the state
space into a number of segments and restrict the set of allowable transactions
within each segment in such a way that the object is invariant-confluent
\emph{within each segment} (even though it may not be globally
invariant-confluent). Then, servers can run coordination-free within a segment
and need only coordinate when transitioning from one segment to another. We now
formalize segmented invariant-confluence, describe the system model we use to
replicate segmented invariant-confluent objects, and introduce a segmented
invariant-confluence decision procedure.

\subsection{Formalism}
Consider a distributed object $O = (S, \join)$, a start state $s_0 \in S$, a
set of transitions $T$, and an invariant $I$. A segmentation $\Sigma = (I_1,
T_1), \ldots, (I_n, T_n)$ is a sequence of $n$ segments $(I_i, T_i)$ where
every $T_i$ is a subset of $T$. $O$ is \defword{segmented invariant-confluent}
with respect to $s_0$, $T$, $I$, and $\Sigma$, abbreviated
\defword{\sTISconfluent}, if the following conditions hold:
\begin{itemize}
  \item
    The start state satisfies the invariant (i.e. $I(s_0)$).

@@

  \item
    $O$ is invariant-confluent within each segment. That is, for every $(I_i,
    T_i) \in \Sigma$ and for every state $s \in I_i$, $O$ is
    \sticonfluent{s}{T_i}{I_i}.
\end{itemize}

@@

\subsection{System Model}
Now, we describe the system model used to replicate a segmented
invariant-confluent object without any coordination within a segment and with
only a small amount of coordination when transitioning between segments. As
before, we replicate an object $O$ across a set $p_1, \ldots, p_n$ of $n$
servers each of which manages a replica $s_i \in S$ of the object. Every server
begins with $s_0$, $T$, $I$, and $\Sigma$. Moreover, at any given point in
time, a server designates one of the segments in $\Sigma$ as its
\defword{active segment}.  Initially, every server chooses the first segment
$(I_i, T_i) \in \Sigma$ such that $I_i(s_0)$ to be its active segment. We'll
see momentarily the significance of the active segment.

As before, servers repeatedly perform one of two actions: execute a transaction
or merge states with another server. Merging states in the segmented
invariant-confluence system model is identical to merging states in the
invariant-confluence system model. A server $p_i$ sends its state $s_i$ to
another server $p_j$ which \emph{must} merge $s_i$ into its state $s_j$.
%
Transaction execution in the new system model, on the other hand, is a bit more
involved. Consider a server $s_i$ with active segment $(I_i, T_i)$. A client
can request that $p_i$ execute a transaction $t \in T$. We consider what
happens when $t \in T_i$ and $t \notin T_i$ separately.

@@

\begin{algorithm}[t]
  \caption{Transaction execution in segmented invariant-confluence system model}%
  \algolabel{TxnExecution}
  \begin{algorithmic}
    \If{$t \notin T_i$}
      \State Execute $t$ with global coordination
    \Else{}
      \If{$I_i(t(s_i))$} Commit $t$
      \ElsIf{$\lnot I(t(s_i))$} Abort $t$
      \Else{} Execute $t$ with global coordination
      \EndIf{}
    \EndIf{}
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Interactive Decision Procedure}
In order to determine whether or not an object $O$ is \sTISconfluent{} we have
to determine whether or not $O$ is invariant-confluent within each segment
$(I_i, T_i) \in \Sigma$. That is, we have determine if $O$ is
\sticonfluent{s}{T_i}{I_i} confluent for every state $s \in I_i$. Ideally, we
could leverage \algoref{InteractiveDecisionProcedure}, invoking it once per
segment. Unfortunately, \algoref{InteractiveDecisionProcedure} can only be used
to determine if $O$ is \sticonfluent{s}{T_i}{I_i} for a \emph{particular} state
$s \in I_i$, not for \emph{every} state $s \in I_i$. Thus, we would have to
invoke \algoref{InteractiveDecisionProcedure} $|I_i|$ times fore every segment
$(I_i, T_I)$, which is clearly infeasible given that $I_i$ can be large or even
infinite.

Instead, we develop a new decision procedure that can be used to determine if
an object is \sticonfluent{s}{T}{I} for every state $s \in I$. To do so, we
need to extend the notion of reachability to a notion of coreachability and
then generalize \thmref{IclosureEquivalentIconfluence}. Two states $s_1, s_2
\in I$ are \defword{coreachable} with respect to $T$ and $I$, abbreviated
\defword{$\TIcoreachable{s_1, s_2}$}, if there exists some state $s_0 \in I$
such that $s_1$ and $s_2$ are both $\sTIreachable{}$.

\begin{theorem}\thmlabel{ExtendedIclosureEquivalentIconfluence}
  Consider an object $O = (S, \join)$, a set of transactions $T$, and an
  invariant $I$. If every pair of states in the invariant are coreachable (i.e.
  $\forall s_1, s_2 \in S.\ \TIcoreachable{s_1, s_2}$), then
  \[
    \text{$O$ is \Iclosed{}}
    \iff
    \text{$O$ is \sticonfluent{s}{T}{I} for every $s \in I$}
  \]
\end{theorem}
\begin{proof}
  The proof of the forward direction is exactly the same as the proof of
  \thmref{IclosureImpliesIconfluence}. Transactions always maintain the
  invariant, so if merge does as well, then every reachable state must satisfy
  the invariant.
  %
  For the reverse direction, consider two arbitrary states $s_1, s_2 \in I$.
  The two points are $\TIcoreachable{}$, so there exists some state $s_0$ from
  which they can be reached. $O$ is \sTIconfluent{} and $s_1 \join s_2$ is
  $\sTIreachable{}$, so it satisfies the invariant.
\end{proof}

Using \thmref{ExtendedIclosureEquivalentIconfluence}, we develop
\algoref{ArbitraryStartInteractiveDecisionProcedure}: a natural generalization
of \algoref{InteractiveDecisionProcedure}.
\algoref{InteractiveDecisionProcedure} iteratively refines the set of
\emph{reachable} states whereas
\algoref{ArbitraryStartInteractiveDecisionProcedure} iteratively refines the
set of \emph{coreachable} states, but otherwise, the core of the two algorithms
is the same.\footnote{%
  Another small difference is that \IsIclosed{} behaves differently in
  \algoref{InteractiveDecisionProcedure} and
  \algoref{ArbitraryStartInteractiveDecisionProcedure}. In
  \algoref{ArbitraryStartInteractiveDecisionProcedure}, \IsIclosed{} returns a
  triple (closed, $s_1$, $s_2$) where if closed is false, then $s_1, s_2 \in I$
  are two states not in $NR$ such that $I(s_1)$ and $I(s_2)$ but $\lnot I(s_1
  \join s_2)$. If no such states exist, then closed is true, and $s_1$ and
  $s_2$ are null.
}
Now, a segmented invariant-confluence decision procedure, can simply invoke
\algoref{ArbitraryStartInteractiveDecisionProcedure} once on each segment.

\newcommand{\algocomment}[1]{\State \textcolor{flatdenim}{\texttt{//} #1}}
\begin{algorithm}[t]
  \caption{%
    Interactive invariant-confluence decision procedure for arbitrary start
    state $s \in I$
  }%
  \algolabel{ArbitraryStartInteractiveDecisionProcedure}
  \begin{algorithmic}
    \algocomment{Return if $O$ is \sticonfluent{s}{T}{I} for every $s \in I$.}
    \Function{IsInvConfluent}{$O$, $T$, $I$}
      \State \Return \Call{Helper}{$O$, $T$, $I$, $\emptyset$, $\emptyset$}
    \EndFunction

    \State

    \algocomment{$R$ is a set of $\TIcoreachable{}$ states.}
    \algocomment{$NR$ is a set of $\TIcounreachable{}$ states.}
    \Function{Helper}{$O$, $T$, $I$, $R$, $NR$}
      \State closed, $s_1$, $s_2$ $\gets$ \Call{IsIclosed}{$O$, $I$, $NR$}
      \If {closed}
        \State \Return true
      \EndIf
      \State Augment $R$ and $NR$ with random search and user input
      \If{$(s_1, s_2) \in R$}
        \Return false
      \Else\
        \Return \Call{Helper}{$O$, $T$, $I$, $R$, $NR$}
      \EndIf
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{example}\examplelabel{CounreachableExample}
  Let $O = (\ints^3 \times \ints^3, \join)$ be a
  PN-Counter~\cite{shapiro2011comprehensive} replicated on $3$ machines. Every
  state $s = (p_1, p_2, p_3), (n_1, n_2, n_3)$ represents the integer $(p_1 +
  p_2 + p_3) - (n_1 + n_2 + n_3)$. To increment and decrement the counter, the
  $i$th server increments $p_i$ and $n_i$ respectively, and $\join$ computes an
  element-wise maximum. Our start state $s_0 = (0, 0, 0), (0, 0, 0)$; our set
  $T$ of transactions consists of increment and decrement; and our invariant
  $I$ is that the value of $s$ is non-negative.

  Applying \algoref{InteractiveDecisionProcedure}, \IsIclosed{} returns false
  with the states $s_1 = (1, 0, 0), (0, 1, 0)$ and $s_2 = (1, 0, 0), (0, 0,
  1)$. Both are reachable, so $O$ is not \sTIconfluent{} and
  \algoref{InteractiveDecisionProcedure} returns false.  The culprit is
  concurrent decrements, which we can forbid in a simple one-segment
  segmentation $\Sigma = (I, T^+)$ where $T^+$ consists only of increment
  transactions. Applying, \algoref{ArbitraryStartInteractiveDecisionProcedure},
  \IsIclosed{} again returns false with states $s_1$ and $s_2$. This time,
  however, the user recognizes that the two states are not
  $\ticoreachable{T^+}{I}{}$. The user refines $NR$ with the observation that
  two states are coreachable if and only if they have the same values of $n_1,
  n_2, n_3$. After this, \IsIclosed{} and
  \algoref{ArbitraryStartInteractiveDecisionProcedure} return true.
\end{example}

\subsection{Discussion}
There are a couple of things worth noting about segmented invariant-confluence,
its system model, and its decision procedure. First, invariant-confluence is a
very coarse-grained property. If an object is invariant-confluent, then we can
replicate it with no coordination. If it is not invariant-confluent, then we
have no guarantees. There's no in-between. Segmented invariant-confluence, on
the other hand, is a much more fine-grained property that can be applied to
applications with varying degrees of complexity. Segmented
invariant-confluence provides guarantees to complex applications that require a
large number of segments and to simple applications with a smaller number of
segments, whereas invariant-confluence only provides guarantees to applications
that can be segmented into a single segment.

Second, segmented invariant-confluence naturally subsumes a distributed locking
approach to replicating non-invariant-confluent objects.  This approach first
recognizes which transactions cannot be safely executed concurrently and then
requires them to acquire a distributed lock before
executing~\cite{balegas2015putting, gotsman2016cause}. For example, in a
banking application with the invariant that all balances remain non-negative,
concurrent deposits are permitted, but concurrent withdrawals can lead to
invariant violations. Thus, we require that withdrawals acquire a distributed
lock before executing. This example is exactly the same as
\exampleref{CounreachableExample} which we handled by simply removing
withdrawal transactions from our segmentation's set of transactions.

@@

Fourth, a segmented invariant-confluence decision procedure can also leverage
\thmref{LatticeProperty} in addition to
\algoref{ArbitraryStartInteractiveDecisionProcedure}. If an object $O$ meets
criteria (1) - (3), then it is \sticonfluent{s}{T}{I} for every state $s \in
I$.

\todo{Forward reference examples in the evaluation section.}
